---
title: "POMA: An R Package for Statistical Analysis of Metabolomic Data"
author: "Pol Castellano-Escuder"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    toc: true
    theme: cayman
    highlight: github
  pdf_document:
    toc: true
---

# Installation

```{r, eval = F}
devtools::install_github("pcastellanoescuder/POMA", dependencies = TRUE)
```

# Load Packages

```{r, message=FALSE, warning=FALSE, comment=FALSE}
library(POMA)
library(tidyverse)
# library(mixOmics)
# library(limma)
# library(impute)
# library(reshape2)
```

# Load Example Data

```{r, message=FALSE, warning=FALSE, comment=FALSE}
data <- vroom::vroom("MET_CRC_ST000284.csv", delim = ",")
covariates <- vroom::vroom("COV_CRC_ST000284.csv", delim = ",")

# Generate 10% of NAs to test PomaImpute() function

set.seed(6)
df <- data[, 3:ncol(data)]
while(sum(is.na(df) == TRUE) < (nrow(df) * ncol(df) * 10/100)){
  df[sample(nrow(df),1), sample(ncol(df),1)] <- NA
}
data <- cbind(data[, 1:2], df)
```

# Imputation

Imputation using KNN method.

```{r}
imputed <- PomaImpute(data, ZerosAsNA = F, RemoveNA = T, cutoff = 20, method = "knn")
```

# Normalization

Generation of 3 different matrix using 3 different normalization methods.

```{r}
not_norm <- PomaNorm(imputed, method = "none", round = 3)
log_pareto <- PomaNorm(imputed, method = "log_pareto", round = 3)
log_transformed <- PomaNorm(imputed, method = "log_transformation", round = 3)
```

## By subjects

Boxplot comparisson between not normalized data and normalized data (using "log_pareto").

```{r, fig.align='center', dpi=200}
PomaNormPlot(not_norm, group = "subjects") + ggtitle("Not Normalized data")
```

```{r, fig.align='center', dpi=200}
PomaNormPlot(log_pareto, group = "subjects") + ggtitle("Normalized data")
```

## By metabolites

Boxplot comparisson between not normalized data and normalized data (using "log_pareto") groupping by metabolites.

```{r, fig.align='center', dpi=200}
PomaNormPlot(not_norm, group = "metabolites") + ggtitle("Not Normalized data")
```

```{r, fig.align='center', dpi=200}
PomaNormPlot(log_pareto, group = "metabolites") + ggtitle("Normalized data")
```

# Univariate analysis

## T-test

```{r}
univ_ttest_not_norm <- PomaUnivariate(not_norm, method = "ttest", adjust = "fdr")
knitr::kable(head(univ_ttest_not_norm))
```

A volcano plot of the results:

```{r, fig.align='center', dpi=200, message=FALSE, warning=FALSE, comment=FALSE}
PomaVolcano(univ_ttest_not_norm, pval = "raw", FC_cutoff = 1.5)
```

## U Mann-Whitney

```{r}
univ_mann_not_norm <- PomaUnivariate(not_norm, method = "mann", adjust = "fdr")
knitr::kable(head(univ_mann_not_norm))
```

A volcano plot of the results:

```{r, fig.align='center', dpi=200, message=FALSE, warning=FALSE, comment=FALSE}
PomaVolcano(univ_mann_not_norm, pval = "raw", FC_cutoff = 1.5)
```

# Multivariate analysis

## sPLS-DA (Supervised ultivariate analysis)

Sparse partial least squares discriminant analysis of log pareto normalized data.

```{r, fig.align='center', dpi=200, message=FALSE, warning=FALSE, comment=FALSE}
multiv_splsda <- PomaMultivariate(log_pareto, method = c("splsda"), components = 5,
                                  validation = "Mfold", folds = 5, nrepeat = 10, 
                                  num_features = 10)

multiv_splsda$splsda_scores_plot
multiv_splsda$bal_error_rate
knitr::kable(head(multiv_splsda$scores_splsda))
knitr::kable(head(multiv_splsda$selected_variables))
```

## PCA (Unsupervised ultivariate analysis)

Principal component analysis of log pareto normalized data.

```{r, fig.align='center', dpi=200, message=FALSE, warning=FALSE, comment=FALSE}
multiv_pca <- PomaMultivariate(log_pareto, method = "pca", components = 4)

multiv_pca$scoresplot
multiv_pca$screeplot
knitr::kable(head(multiv_pca$score_data))
```

# Limma

## Limma without covariates

```{r, fig.align='center', dpi=200, message=FALSE, warning=FALSE, comment=FALSE}
limma_res <- PomaLimma(log_transformed, contrast = "C-H", adjust = "fdr")

limma_res %>%
  rownames_to_column("Metabolite") %>%
  mutate(Fold_Change_Ratio = 2^logFC) %>%
  column_to_rownames("Metabolite") %>%
  PomaVolcano(pval = "raw", FC_cutoff = 1.2, xlim = 1) + ggtitle("Comparisson: Group1/Group2")

knitr::kable(head(limma_res))
```

## Limma with covariates

```{r, fig.align='center', dpi=200, message=FALSE, warning=FALSE, comment=FALSE}
limma_res2 <- PomaLimma(log_transformed, covariates = covariates, contrast = "C-H", adjust = "fdr")

limma_res2 %>%
  rownames_to_column("Metabolite") %>%
  mutate(Fold_Change_Ratio = 2^logFC) %>%
  column_to_rownames("Metabolite") %>%
  PomaVolcano(pval = "raw", FC_cutoff = 1.2, xlim = 1) + ggtitle("Comparisson: Group1/Group2")

knitr::kable(head(limma_res2))
```

